Shooting Occurrences in Toronto have increased substantially all throughout the city: A Data Analysis

### 2022-02-05 Shooting_Occurrences/outputs/paper/Shooting_Occurrences.Rmd
Final edit

---
title: "Shooting Occurrences in Toronto have increased substantially all throughout the city: A Data Analysis"
author: 
  - Thomas D'Onofrio
thanks: "Code and data are available at: https://github.com/TDonofrio62/Paper-on-Shooting-Occurrences-in-Toronto."
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "Toronto, which was once seen as a safe place to live, has seemingly had a lot more gun violence occur within the city over the years. Data on shooting occurrences in Toronto, including the years and police divisions in which they occurred ,has been obtained, and will be analyzed in this paper with the use of several figures. It was found that shooting occurrences have greatly increased all throughtout the city of Toronto, and overall have more than doubled since 2014. Our discovery unfortunately has some implications, as police have racial biases against people of colour which have caused the dataset worked with to be biased."
output:
  bookdown::pdf_document2
toc: FALSE
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(knitr)
library(opendatatoronto)
```

@@

The dataset being used in this paper was taken from the City of Toronto's open data portal using the opendata toronto package [@citeODT]. It is a subset of a larger collection of data called The Toronto's Police Services Annual Statistical Report. This report contains data based on police related topics including a variety of different types of crimes and other incidents, and administrative information. The dataset that will be analyzed draws from the "Shooting Occurrences" portion of the annual statistical report. It contains all Toronto shooting incidents between 2014 and 2019. A shooting occurrence is described as an altercation where a projectile was shot from a firearm and injured a person [@citeTPS]. Toronto Police Services do not count suicidal attempts or shots taken by Police in the data.

@@

shooting_occurrences <- read_csv(here::here("inputs/data/shooting_occurences.csv"))

shooting_occurrences %>% 
  slice(1,2,18,19,36,37,54,55,72,73,90,91) %>%
  select(occurred_year, geo_division, num_shootings) %>%
  kable(
    caption= "12 Observations from dataset of shooting occurences in Toronto",
    col.names = c("Year", "Geographic Division", "Number of Shootings"),
    booktabs = TRUE,
    linesep = "",
    align = c('c', 'c', 'c')
  )

@@

Occurred year is fairly self explanatory, and is the year in which all the shootings in an observation took place. It is a numerical variable, but was also converted to a categorical at points for use in visulaizations seen later. Geo division is a categorical variable representing the geographically determined area boundaries for each police division. Num of shootings is a numerical variable that counts the number of shootings corresponding to each observation.

@@

```{r yearlyshootings, echo = FALSE}
shooting_occurrences %>%
  select(occurred_year,num_shootings)%>%
  group_by(occurred_year)%>%
  summarize(number_of_shootings = sum(num_shootings)) %>%
  kable(
    caption= "Number of Shootings in Toronto each year from 2014-2019",
    col.names = c("Year", "Number of Shootings"),
    booktabs = TRUE,
    linesep = "",
    align = c('c', 'c')
  )

@@

```{r scatter, fig.cap= "Relationship between number of shootings in each police division and year", echo = FALSE, fig.height=4.1}
shooting_occurrences %>%
  ggplot(aes(x = occurred_year, y = num_shootings, color = geo_division)) +
  geom_point()+
  theme_bw() +
  theme(legend.key.size = unit(0.5, 'cm'))+
  labs(x = "Year",
       y = "Number of Shootings",
       color = "Geographic Division"
       )
```

@@

```{r bar1, fig.cap= "Bar plot showing number of shootings in each police division for every year", echo = FALSE, fig.height=4.1, fig.pos = "h"}
shooting_occurrences %>%
  ggplot(aes(x = occurred_year, y = num_shootings, fill = geo_division)) + 
  geom_bar(stat='identity') +
  theme_bw() +
  theme(legend.key.size = unit(0.5, 'cm'))+
  labs(x = "Year",
       y = "Number of Shootings",
       fill = "Geographic Division"
       )+
  scale_fill_viridis_d(option="turbo")+
  scale_x_continuous(breaks = c(2014,2015,2016,2017,2018,2019)) 
```

@@

Figure \@ref(fig:bar1) also makes the changes in number of shootings in each police division more evident. Every single coloured section representing the divisions are larger in 2019 than 2014, and the number of shootings for most divisons increased every single year. The larger police divisions specifically saw massive growths in the size of their colour in the bars of the Figure Figure \@ref(fig:bar1), which was also, albeit slightly less, clear in Figure \@ref(fig:scatter). This can be seen in D12's, D23's, and D31's coloured parts of the bars, as their filled in areas in the bars are very large in 2019 compared to what they were in 2014. A final figure, Figure \@ref(fig:bar2), can be used to show how the number of shootings has changed in each division even further.

```{r bar2, fig.cap= "Bar plot showing number of shootings every year for each police divison", echo = FALSE, fig.height=4, fig.width=8, fig.pos = "h"}
shooting_occurrences %>%
  ggplot(aes(x = geo_division, y = num_shootings, 
             fill = as.character(occurred_year))) + 
  geom_bar(stat='identity', position = position_stack(reverse = TRUE)) + 
  theme_bw() +
  labs(x = "Geographic Division",
       y = "Number of Shootings",
       fill = "Year"
       )+
  scale_fill_viridis_d()
```

@@

Overall, our data has painted a clear story about the shooting occurences in Toronto. It is extremely evident based on Table \@ref(tab:yearlyshootings), Figure \@ref(fig:scatter), and Figure \@ref(fig:bar1) that the number of shooting occurrences in Toronto have increased greatly from 2014 to 2019, and that this increase has been occurring steadily over the years since 2014. While the other figures helped, Figure \@ref(fig:bar2) made it clear that shooting occurrences have been increasing all throughout Toronto in every single police division, although some of the larger divisions in particular have seen even bigger increases than the rest. It is still important to note that the racial bias towards people of colour by the Toronto Police has caused our dataset to be biased, which may have caused implications for our findings.

>>>>

---
title: "Shooting Occurrences in Toronto have increased substantially all throughout the city: A Data Analysis"
author: 
  - Thomas D'Onofrio
thanks: "Code and data are available at: https://github.com/TDonofrio62/Shooting-Occurrences-in-Toronto"
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "Toronto, which was once seen as a safe place to live, has seemingly had a lot more gun violence occur within the city over the years. Data on shooting occurrences in Toronto, including the years and police divisions in which they occurred ,has been obtained, and will be analyzed in this paper with the use of several figures. It was found that shooting occurrences have greatly increased all throughtout the city of Toronto, and overall have more than doubled since 2014. Our discovery unfortunately has some implications, as police have racial biases against people of colour which have caused the dataset worked with to be biased."
output:
  bookdown::pdf_document2
toc: FALSE
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
#Load libraries
library(tidyverse)
library(knitr)
library(opendatatoronto)
```

@@

The dataset being used in this paper was taken from the City of Toronto's open data portal using the open data Toronto package [@citeODT]. It is a subset of a larger collection of data called The Toronto's Police Services Annual Statistical Report. This report contains data based on police related topics including a variety of different types of crimes and other incidents, and administrative information. The dataset that will be analyzed draws from the "Shooting Occurrences" portion of the annual statistical report. It contains all Toronto shooting incidents between 2014 and 2019. A shooting occurrence is described as an altercation where a projectile was shot from a firearm and injured a person [@citeTPS]. Toronto Police Services do not count suicidal attempts or shots taken by Police in the data.

@@

#Load in dataset
shooting_occurrences <- read_csv(here::here("inputs/data/shooting_occurences.csv"))

#Show a brief extract of the dataset by selecting 12 observations in a table
shooting_occurrences %>% 
  slice(1,2,18,19,36,37,54,55,72,73,90,91) %>%
  select(occurred_year, geo_division, num_shootings) %>%
  kable(
    caption= "12 Observations from dataset of shooting occurences in Toronto",
    col.names = c("Year", "Geographic Division", "Number of Shootings"),
    booktabs = TRUE,
    linesep = "",
    align = c('c', 'c', 'c')
  )

@@

Occurred year is fairly self explanatory, and is the year in which all the shootings in an observation took place. It is a numerical variable, but was also converted to a categorical at points for use in visualizations seen later. Geo division is a categorical variable representing the geographically determined area boundaries for each police division. Num of shootings is a numerical variable that counts the number of shootings corresponding to each observation.

@@

```{r yearlyshootings, echo = FALSE}
#Create a table that compares the number of shooting occurrences and the year in which they occurred
#Use group_by function to group all the shootings in the same year together and show how many occurred
shooting_occurrences %>%
  select(occurred_year,num_shootings)%>%
  group_by(occurred_year)%>% 
  summarize(number_of_shootings = sum(num_shootings)) %>%
  kable(
    caption= "Number of Shootings in Toronto each year from 2014-2019",
    col.names = c("Year", "Number of Shootings"),
    booktabs = TRUE,
    linesep = "",
    align = c('c', 'c')
  )

@@

```{r scatter, fig.cap= "Relationship between number of shootings in each police division and year", echo = FALSE, fig.height=4.1}
#Create a scatterplot to show each observation in the dataset, comparing the number of shootings to the year in which they occurred
#Colour the points based on the police division where the group of shootings occurred
shooting_occurrences %>%
  ggplot(aes(x = occurred_year, y = num_shootings, colour = geo_division)) +
  geom_point()+
  theme_bw() +
  theme(legend.key.size = unit(0.5, 'cm'))+
  labs(x = "Year",
       y = "Number of Shootings",
       color = "Geographic Division"
       )
```

@@

```{r bar1, fig.cap= "Bar plot showing number of shootings in each police division for every year", echo = FALSE, fig.height=4.1, fig.pos = "h"}
#Create a bar plot to compare the number of shootings to the year in which they occurred
#Make each bar be broken up into coloured sections, using fill, to represent the police divisions in which a certain amount of shootings occurred each year
shooting_occurrences %>%
  ggplot(aes(x = occurred_year, y = num_shootings, fill = geo_division)) + 
  geom_bar(stat='identity') +
  theme_bw() +
  theme(legend.key.size = unit(0.5, 'cm'))+
  labs(x = "Year",
       y = "Number of Shootings",
       fill = "Geographic Division"
       )+
  scale_fill_viridis_d(option="turbo")+
  scale_x_continuous(breaks = c(2014,2015,2016,2017,2018,2019)) 
```

@@

Figure \@ref(fig:bar1) also makes the changes in number of shootings in each police division more evident. Every single coloured section representing the divisions are larger in 2019 than 2014, and the number of shootings for most divisions increased every single year. The larger police divisions specifically saw massive growths in the size of their colour in the bars of the Figure Figure \@ref(fig:bar1), which was also, albeit slightly less, clear in Figure \@ref(fig:scatter). This can be seen in D12's, D23's, and D31's coloured parts of the bars, as their filled in areas in the bars are very large in 2019 compared to what they were in 2014. A final figure, Figure \@ref(fig:bar2), can be used to show how the number of shootings has changed in each division even further.

```{r bar2, fig.cap= "Bar plot showing number of shootings every year for each police divison", echo = FALSE, fig.height=4, fig.width=8, fig.pos = "h"}
#Create a bar plot to compare the number of shootings to the police division where they occurred
#Make each bar be broken up into coloured sections, using fill, to represent the year in which a certain amount of shootings occurred within each police division
shooting_occurrences %>%
  ggplot(aes(x = geo_division, y = num_shootings, 
             fill = as.character(occurred_year))) + 
  geom_bar(stat='identity', position = position_stack(reverse = TRUE)) + 
  theme_bw() +
  labs(x = "Geographic Division",
       y = "Number of Shootings",
       fill = "Year"
       )+
  scale_fill_viridis_d()
```

@@

Overall, our data has painted a clear story about the shooting occurrences in Toronto. It is extremely evident based on Table \@ref(tab:yearlyshootings), Figure \@ref(fig:scatter), and Figure \@ref(fig:bar1) that the number of shooting occurrences in Toronto have increased greatly from 2014 to 2019, and that this increase has been occurring steadily over the years since 2014. While the other figures helped, Figure \@ref(fig:bar2) made it clear that shooting occurrences have been increasing all throughout Toronto in every single police division, although some of the larger divisions in particular have seen even bigger increases than the rest. It is still important to note that the racial bias towards people of colour by the Toronto Police has caused our dataset to be biased, which may have caused implications for our findings.
